---
title: "Homework 1: Namit Shrivastava"
format: pdf
editor: visual
---
Firstly I will be creating the dataset
```{r}
film_noir_data <- data.frame(
  Week = as.Date(c("2022-10-02", "2022-10-09", "2022-10-16", "2022-10-23", 
                   "2022-10-30", "2022-11-06", "2022-11-13", "2022-11-20",
                   "2022-11-27", "2022-12-04", "2022-12-11", "2022-12-18")),
  Searches = c(68, 73, 58, 59, 72, 70, 77, 57, 56, 76, 63, 52)
)
```
## 1. a) [5 points] Calculate maximum likelihood estimate of p (i.e. the proportion of all 781 searches that occurred in each week). Graph these 12 proportions.

Ok so by formula maximum likelihood estimate (MLE) of p for each week is simply the proportion of searches in that week (each row) relative to the total searches i.e. 
Pi = (searches in week i)/total searches
For example, in the table, for Week 10/2/2022, the p value will be calculated as 68/781 = 0.0871 and so on for the other rows. The results are:
```{r}
#Now I will be calculating the 
#proportions with formatted strings
total_searches <- sum(film_noir_data$Searches)
film_noir_data$Proportion <- sprintf("%d/%d = %.4f", 
                                   film_noir_data$Searches, 
                                   total_searches,
                                   film_noir_data$Searches/total_searches)

film_noir_data
```

