---
title: "Homework 2: Namit Shrivastava"
format: pdf
editor: visual
---
## 1. [5 points] A researcher who wants to study the predictors of lung cancer draws a sample of persons with lung cancer from a database of lung cancer patients and a sample of persons without lung cancer from the general population. Is this a prospective or retrospective study?

I think this is a retrospective study. 
Now from the class lecture, in a retrospective study, researchers look back in time to examine data or events that have already occurred. So in this scenario, the researcher is selecting a group of individuals who already have lung cancer (cases) and comparing them to a group of individuals without lung cancer (controls). Simply the goal here is to investigate potential predictors or risk factors by analyzing past exposures or characteristics in these two groups.

So the key aspect to be pointed here is that the researcher is not following participants over time to see who develops lung cancer (which in turn would make it a prospective study). Instead, they are starting with an outcome (lung cancer) that has already happened and working backward to identify possible causes or predictors. This backward-looking approach in the question mentioned is what defines a retrospective study.


## 2. [5 Points] A researcher who wants to study the impact of Vitamin D on children’s health draws a sample of children, randomly splits them into two groups and gives one group Vitamin D and the other group a placebo. Is this a prospective or retrospective study?

I believe this is a prospective study.
In this scenario, the researcher is starting with a group of children who are initially healthy and then actively intervening by giving one group Vitamin D and the other group a placebo. The researcher will follow these groups over time to observe how their health outcomes differ. This forward-looking approach i.e. starting from the intervention and moving forward to measure its effects, is characteristic of a prospective study as discussed in the lecture.

Also I would like to point out that what makes it even clearer is that the study involves randomization, which is a hallmark of experimental designs like randomized controlled trials. These types of studies are inherently prospective because they track participants into the future to assess the impact of the intervention. So in this case, Vitamin D on specific outcomes i.e. children’s health. So, I’m confident this would be classified as a prospective study.

## 3. The following data are based on a study (Petrovčič, et al, 2016) that varied the content of an email message asking persons to participate in a survey. One group received a message that included a “plea for help.” The other group received a message that did NOT include a plea for help. Here are the results:
| Message includes a plea for help | Yes | No |
|:--------------------------------|:---:|:---:|
| Yes                             | 117 | 1,131 |
| No                              | 94  | 1,158 |



## 3. a) [10 points] Estimate the relative risk of responding (plea for help vs not), and report a 95% confidence interval for log-relative risk.

Let me define the variables:
* $n_{11} = 117$ (Yes responses with plea)
* $n_{12} = 1131$ (No responses with plea)
* $n_{21} = 94$ (Yes responses without plea)
* $n_{22} = 1158$ (No responses without plea)
* $n_{1+} = n_{11} + n_{12} = 1248$ (Total with plea)
* $n_{2+} = n_{21} + n_{22} = 1252$ (Total without plea)

1) Calculating risks for each group:

Risk (plea for help) = $p_1 = \frac{n_{11}}{n_{1+}} = \frac{117}{1248} = 0.0938$ or 9.38%  
Risk (no plea) = $p_2 = \frac{n_{21}}{n_{2+}} = \frac{94}{1252} = 0.0751$ or 7.51%

2) The relative risk:
$RR = \frac{p_1}{p_2} = \frac{0.0938}{0.0751} = 1.249$

3) For the 95% CI of log(RR):
$log(RR) \pm 1.96 \times SE(log(RR))$

Where $SE(log(RR)) = \sqrt{\frac{1-p_1}{n_{1+}p_1} + \frac{1-p_2}{n_{2+}p_2}}$

$SE(log(RR)) = \sqrt{\frac{1-0.0938}{1248(0.0938)} + \frac{1-0.0751}{1252(0.0751)}} = 0.1324$

95% CI for log(RR): $ln(1.249) \pm 1.96(0.1324) = 0.223 \pm 0.259 = (-0.0365, 0.4825)$

Therefore, 95% CI for RR: $(e^{-0.0365}, e^{0.4825}) = (0.963, 1.619)$

Now, let me verify this using R:
```{r}
# Data from the table
yes_plea <- 117
no_plea <- 1131 
yes_no_plea <- 94
no_no_plea <- 1158

# Risks
risk1 <- yes_plea / (yes_plea + no_plea)
risk2 <- yes_no_plea / (yes_no_plea + no_no_plea)

# Relative risk
rr <- risk1 / risk2

# Log RR and SE
log_rr <- log(rr)
# Corrected SE formula
se_log_rr <- sqrt((1-risk1)/(risk1*(yes_plea + no_plea)) + 
                  (1-risk2)/(risk2*(yes_no_plea + no_no_plea)))

# 95% CI using dynamic z value function
z <- qnorm(0.975)
z
ci_lower <- log_rr - z * se_log_rr
ci_upper <- log_rr + z * se_log_rr

# Results
cat("Relative Risk:", round(rr, 3), "\n")
cat("95% CI:", round(ci_lower, 3), "-", round(ci_upper, 3), "\n")
```

## 3b) [10 points] Estimate the odds ratio (plea for help vs not), and report a 95% confidence interval for the log- odds ratio.

Let me define the variables again:
* $n_{11} = 117$ (Yes responses with plea)
* $n_{12} = 1131$ (No responses with plea)
* $n_{21} = 94$ (Yes responses without plea)
* $n_{22} = 1158$ (No responses without plea)

1) Calculating odds for each group:

Odds (plea for help) = $\frac{n_{11}}{n_{12}} = \frac{117}{1131} = 0.1034$
Odds (no plea) = $\frac{n_{21}}{n_{22}} = \frac{94}{1158} = 0.0812$

2) The odds ratio:
$OR = \frac{n_{11}n_{22}}{n_{12}n_{21}} = \frac{117 \times 1158}{1131 \times 94} = 1.274$

3) For 95% CI of log(OR):
$SE(log(OR)) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$

$SE(log(OR)) = \sqrt{\frac{1}{117} + \frac{1}{1131} + \frac{1}{94} + \frac{1}{1158}} = 0.14468$

95% CI for log(OR): $ln(1.274) \pm 1.96(0.14468) = 0.242 \pm 0.274 = (-0.0418, 0.526)$

Therefore, 95% CI for OR: $(e^{-0.0318}, e^{0.5158}) = (0.959, 1.692)$

So the code for this calculation is:
```{r}
# Defining the counts
a <- 117  # responded with plea
b <- 1131 # did NOT respond with plea
c <- 94   # responded without plea
d <- 1158 # did NOT respond without plea

# Calculate odds for each group
odds_plea <- a / b
odds_no_plea <- c / d

# Calculate odds ratio
odds_ratio <- odds_plea / odds_no_plea

# Log-odds ratio and standard error
log_or <- log(odds_ratio)
se_log_or <- sqrt((1 / a) + (1 / b) + (1 / c) + (1 / d))
se_log_or

# 95% Confidence interval for the log-odds ratio
z <- qnorm(0.975)
z
ci_lower_log <- log_or - z * se_log_or
ci_upper_log <- log_or + z * se_log_or

# Print results
cat("Odds Ratio:", odds_ratio, "\n")
cat("95% CI for Odds Ratio: (", ci_lower_log, ",", ci_upper_log, ")\n")
```

## 3c) [5 points] Summarize and interpret your findings from parts a) and b). Does the “plea for help” improve response rates?

Ok so in part 3a) the relative risk (RR) of responding to a message with a “plea for help” compared to a message without a plea is 1.249. This means that the risk of responding is about 24.9% higher for those who received the plea for help. However, the 95% confidence interval (CI) for the relative risk is (−0.038, 0.482). Since this range includes 0, suggesting that the observed effect is not statistically significant. The interval implies that one cannot rule out the possibility that there is no effect or even a negative effect of the plea for help on response rates.

Now in part 3b) The odds ratio (OR) of responding to a message with a “plea for help” versus without a plea is 1.274. This indicates that the odds of responding are approximately 27.4% higher with the plea for help. The 95% confidence interval for the log-odds ratio is (−0.041, 0.526). Like the relative risk interval, this range also includes 0, indicating no statistically significant effect. This means that while there is a positive trend, one cannot be confident that the plea for help truly increases the odds of response.

Hence, from my analysis, it looks like both the relative risk and odds ratio suggest that including a plea for help in the message might boost response rates. However, since the confidence intervals (logs scale) for both measures include 0, one can't say the observed effects are statistically significant. In other words, one can't definitively conclude that the plea for help actually improves response rates based on this data alone. While the plea seems to have a positive effect, the lack of statistical significance means we should be cautious about our conclusions. We might need more data or larger studies to be sure of its impact.

## 4a) [10 points] The following table is loosely based upon a study of the impact of different types of incentives on survey response rates (Deutskens, et al., 2004). Cases were randomized to either receiver a voucher that the respondent could spend at specific online vendors, or a donation would be made on their behalf. The first question is whether vouchers produce lower or higher response rates relative to donations. Calculate the odds ratio of a voucher producing response relative to donation. Calculate the deviance (G2).

| Incentive Type | Yes | No |
|:---------------|:---:|:---:|
| Voucher        | 166 | 564 |
| Donation       | 121 | 609 |


Let me define the variables for this part:
* $n_{11} = 166$ (Yes responses with voucher)
* $n_{12} = 564$ (No responses with voucher)
* $n_{21} = 121$ (Yes responses with donation)
* $n_{22} = 609$ (No responses with donation)
* $n_{1+} = n_{11} + n_{12} = 730$ (Total with voucher)
* $n_{2+} = n_{21} + n_{22} = 730$ (Total with donation)
* $n_{+1} = n_{11} + n_{21} = 287$ (Total Yes responses)
* $n_{+2} = n_{12} + n_{22} = 1173$ (Total No responses)
* $N = n_{1+} + n_{2+} = 1460$ (Total sample size)

1) Calculating odds ratio:

Odds (voucher) = $\frac{n_{11}}{n_{12}} = \frac{166}{564} = 0.294$
Odds (donation) = $\frac{n_{21}}{n_{22}} = \frac{121}{609} = 0.199$

Odds Ratio = $OR = \frac{n_{11}n_{22}}{n_{12}n_{21}} = \frac{166 \times 609}{564 \times 121} = 1.48$

2) Expected frequencies under independence:
* $E_{11} = \frac{n_{1+}n_{+1}}{N} = \frac{730 \times 287}{1460} = 143.5$
* $E_{12} = \frac{n_{1+}n_{+2}}{N} = \frac{730 \times 1173}{1460} = 586.5$
* $E_{21} = \frac{n_{2+}n_{+1}}{N} = \frac{730 \times 287}{1460} = 143.5$
* $E_{22} = \frac{n_{2+}n_{+2}}{N} = \frac{730 \times 1173}{1460} = 586.5$

3) Deviance calculation:
$G^2 = 2\sum_{i} O_i \ln(\frac{O_i}{E_i})$

$G^2 = 2[(166\ln\frac{166}{143.5}) + (564\ln\frac{564}{586.5}) + (121\ln\frac{121}{143.5}) + (609\ln\frac{609}{586.5})]$

$G^2 = 8.81$

Now, let me use R to get these values:

```{r}
# Data from the table
voucher_yes <- 166
voucher_no <- 564
donation_yes <- 121
donation_no <- 609

# Odds Ratio alculation
odds_voucher <- voucher_yes / voucher_no
odds_donation <- donation_yes / donation_no
odds_ratio <- odds_voucher / odds_donation

# Deviance (G2) calculation
observed <- c(voucher_yes, voucher_no, donation_yes, donation_no)

# Calculating expected counts
total_voucher <- voucher_yes + voucher_no
total_donation <- donation_yes + donation_no
total_yes <- voucher_yes + donation_yes
total_no <- voucher_no + donation_no
total_all <- total_voucher + total_donation

expected_voucher_yes <- (total_voucher * total_yes) / total_all
expected_voucher_no <- (total_voucher * total_no) / total_all
expected_donation_yes <- (total_donation * total_yes) / total_all
expected_donation_no <- (total_donation * total_no) / total_all

expected <- c(expected_voucher_yes, expected_voucher_no, 
              expected_donation_yes, expected_donation_no)

# Compute G2
g2 <- 2 * sum(observed * log(observed / expected))

# Print results
cat("Odds Ratio:", round(odds_ratio, 3), "\n")
cat("Deviance (G²):", round(g2, 3), "\n")
```



## 4b) [10 points] Next, we want to compare vouchers to a lottery. Calculate the odds ratio for a voucher to produce response relative to a lottery. Calculate the deviance (G2). 

| Incentive Type | Yes | No |
|:---------------|:---:|:---:|
| Voucher        | 166 | 564 |
| Lottery        | 132 | 598 |


Let me define the variables:
* $n_{11} = 166$ (Yes responses with voucher)
* $n_{12} = 564$ (No responses with voucher)
* $n_{21} = 132$ (Yes responses with lottery)
* $n_{22} = 598$ (No responses with lottery)
* $n_{1+} = n_{11} + n_{12} = 730$ (Total with voucher)
* $n_{2+} = n_{21} + n_{22} = 730$ (Total with lottery)
* $n_{+1} = n_{11} + n_{21} = 298$ (Total Yes responses)
* $n_{+2} = n_{12} + n_{22} = 1162$ (Total No responses)
* $N = n_{1+} + n_{2+} = 1460$ (Total sample size)

1) First, calculating odds ratio:

Odds (voucher) = $\frac{n_{11}}{n_{12}} = \frac{166}{564} = 0.294$
Odds (lottery) = $\frac{n_{21}}{n_{22}} = \frac{132}{598} = 0.221$

Odds Ratio = $OR = \frac{n_{11}n_{22}}{n_{12}n_{21}} = \frac{166 \times 598}{564 \times 132} = 1.33$

2) For the deviance calculation:

Expected frequencies under independence:
$E_{11} = \frac{n_{1+}n_{+1}}{N} = \frac{730 \times 298}{1460} = 149$
$E_{12} = \frac{n_{1+}n_{+2}}{N} = \frac{730 \times 1162}{1460} = 581$
$E_{21} = \frac{n_{2+}n_{+1}}{N} = \frac{730 \times 298}{1460} = 149$
$E_{22} = \frac{n_{2+}n_{+2}}{N} = \frac{730 \times 1162}{1460} = 581$

3) Calculate components of $G^2$:

$G^2 = 2[(166\ln\frac{166}{149}) + (564\ln\frac{564}{581}) + (132\ln\frac{132}{149}) + (598\ln\frac{598}{581})]$

$= 2[166(0.108) + 564(-0.0296) + 132(-0.121) + 598(0.0288)]$

$= 2[17.95 - 16.72 - 15.99 + 17.24]$

$= 2(2.48)$

$= 4.883$

```{r}
# Data from the table
voucher_yes <- 166
voucher_no <- 564
lottery_yes <- 132
lottery_no <- 598

# Odds Ratio calculation
odds_voucher <- voucher_yes / voucher_no
odds_lottery <- lottery_yes / lottery_no
odds_ratio <- odds_voucher / odds_lottery

# Deviance (G2) calculation
observed <- c(voucher_yes, voucher_no, lottery_yes, lottery_no)

# Calculating expected counts
total_voucher <- voucher_yes + voucher_no
total_lottery <- lottery_yes + lottery_no
total_yes <- voucher_yes + lottery_yes
total_no <- voucher_no + lottery_no
total_all <- total_voucher + total_lottery

expected_voucher_yes <- (total_voucher * total_yes) / total_all
expected_voucher_no <- (total_voucher * total_no) / total_all
expected_lottery_yes <- (total_lottery * total_yes) / total_all
expected_lottery_no <- (total_lottery * total_no) / total_all

expected <- c(expected_voucher_yes, expected_voucher_no, 
             expected_lottery_yes, expected_lottery_no)

# Compute G2
g2 <- 2 * sum(observed * log(observed / expected))

# Print results
cat("Odds Ratio:", round(odds_ratio, 3), "\n")
cat("Deviance (G²):", round(g2, 3), "\n")
```


## 4c) [10 points] Describe the results from the analysis of 4a and 4b. Does there appear to be differences in response rates across each of the type of incentive comparisons in 4a and 4b?

4a) (Voucher vs. Donation):
Odds Ratio (OR): 1.481
Deviance (G2): 8.812
The odds ratio of 1.481 tells me that participants offered a voucher had roughly 48% higher odds of responding compared to those offered a donation. The deviance statistic (G2 = 8.812) is well above the critical value of 3.84 (for a chi-square test at alpha = 0.05), which means this difference is statistically significant. In practical terms, vouchers appear to be more effective at driving responses than donations.

4b) (Voucher vs. Lottery):
Odds Ratio (OR): 1.333
Deviance (G2): 4.883
Here, the odds ratio of 1.333 suggests that vouchers still outperform lotteries, but the effect is smaller i.e. participants offered vouchers had about 33% higher odds of responding than those offered a lottery. The deviance (G2 = 4.883) is also statistically significant (since 4.883 > 3.84), but the smaller odds ratio implies a weaker association compared to vouchers vs. donations.

Hence, vouchers consistently produce higher response rates than both donations and lotteries. However, the magnitude of improvement depends on the comparison i.e. vouchers are much better than donations and moderately better than lotteries.

So talking about the practical implications of these results, say if I were advising on incentive strategies, I’d recommend prioritizing vouchers over both donations and lotteries. The stronger effect against donations makes vouchers a particularly compelling choice in that comparison. However, if logistical constraints (e.g., cost) make lotteries more feasible, the smaller but still significant advantage of vouchers might need to be weighed against those practical considerations.
In short, yep, there are differences in response rates, with vouchers being the most effective incentive overall. The strength of this advantage varies depending on what they’re being compared to.


## 4d) [10 points] Returning to the data from 4a. The deviance can tell us about association, but not about the direction of that association. Calculate a 95% confidence interval for the odds ratio calculated in 4a. Based on the odds ratio, which form of the incentive has the higher response rate? Is this difference significant?

Let me define our variables:
* $n_{11} = 166$ (Yes responses with voucher)
* $n_{12} = 564$ (No responses with voucher)
* $n_{21} = 121$ (Yes responses with donation) 
* $n_{22} = 609$ (No responses with donation)
* $n_{1+} = n_{11} + n_{12} = 730$ (Total with voucher)
* $n_{2+} = n_{21} + n_{22} = 730$ (Total with donation)
* $n_{+1} = n_{11} + n_{21} = 287$ (Total Yes responses)
* $n_{+2} = n_{12} + n_{22} = 1173$ (Total No responses)
* $N = n_{1+} + n_{2+} = 1460$ (Total sample size)

1) First, calculating the odds for each group:

Odds (voucher) = $\frac{n_{11}}{n_{12}} = \frac{166}{564} = 0.2943$
Odds (donation) = $\frac{n_{21}}{n_{22}} = \frac{121}{609} = 0.1987$

2) The odds ratio:
$OR = \frac{n_{11}n_{22}}{n_{12}n_{21}} = \frac{166 \times 609}{564 \times 121} = 1.481$

3) Taking natural logarithm:
$ln(OR) = ln(1.481) = 0.393$

4) Standard error calculation:
$SE_{ln(OR)} = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$

$SE_{ln(OR)} = \sqrt{\frac{1}{166} + \frac{1}{564} + \frac{1}{121} + \frac{1}{609}} = \sqrt{0.0177} = 0.133$

5) 95% CI for log(OR):
$0.393 \pm 1.96(0.133) = (0.132, 0.654)$

6) Therefore, 95% CI for OR:
$95\% CI: (e^{0.132}, e^{0.654}) = (1.14, 1.92)$

Now the R code for this will be
```{r}
# Data from 4a: Voucher vs. Donation
voucher_yes <- 166
voucher_no <- 564
donation_yes <- 121
donation_no <- 609

# Calculate odds ratio manually
odds_voucher <- voucher_yes / voucher_no
odds_donation <- donation_yes / donation_no
odds_ratio <- odds_voucher / odds_donation

# Log-odds ratio and standard error
log_or <- log(odds_ratio)
se_log_or <- sqrt((1 / voucher_yes) + (1 / voucher_no) + 
                  (1 / donation_yes) + (1 / donation_no))

# 95% confidence interval for log-odds ratio
z <- qnorm(0.975) # 1.96 for 95% CI
ci_lower_log <- log_or - z * se_log_or
ci_upper_log <- log_or + z * se_log_or

# Convert back to odds ratio scale
ci_lower <- exp(ci_lower_log)
ci_upper <- exp(ci_upper_log)

# Print results
cat("Odds Ratio:", round(odds_ratio, 3), "\n")
cat("95% Confidence Interval: (", round(ci_lower, 3), ",", round(ci_upper, 3), ")\n")
```

Right, so based on the odds ratio of 1.48 (95% CI: 1.14–1.92), the voucher incentive clearly leads to a higher response rate compared to donations. Since the odds ratio is greater than 1, this means participants offered vouchers had 48% higher odds of responding than those offered donations. In simpler terms, vouchers are more effective at encouraging people to complete the survey. 
Crucially, this difference is statistically significant because the confidence interval (1.14–1.92) doesn’t include 1. If there were no real difference, one would expect the odds ratio to hover around 1, but here the entire interval sits above 1, which tells me one can be 95% confident that vouchers genuinely outperform donations.

Interestingly, what makes me even more confident in this conclusion is both the direction and magnitude of the effect. The odds ratio isn’t just barely above 1. It is solidly in a meaningful improvement territory!
Even the lower bound of the interval (1.14) suggests a 14% increase in odds, which isn’t trivial, while the upper bound (1.92) hints that the true effect could be as large as a 92% boost. This isn’t just a minor difference but a substantial advantage.

If I were to design this survey, I’d prioritize vouchers over donations without hesitation. Vouchers aren’t just slightly better but are significantly better at driving responses.
