---
title: "Homework 3: Namit Shrivastava"
format: pdf
editor: visual
---

## 1. The following data come from a case-control study. The cases were sampled from a registry of all lung cancer patients at a set of 6 clinics. The controls were sampled from the patients at the 6 clinics who did not have lung cancer. Each group was asked if they had ever been regular smokers. The researchers made the following claims (1a-1f) based upon these data. State whether the claim is TRUE or FALSE and explain your answer. In this case, the population of interest is those persons who visited the 6 clinics over a specified time period.

| Smoker | Lung Cancer |     
|        | Yes | No    |
|:-------|:---:|:-----:|
| Yes    | 126 | 100   |
| No     | 35  | 61    |

First let me define variables:
* $n_{11} = 126$ (Smokers with lung cancer)
* $n_{12} = 100$ (Smokers without lung cancer)
* $n_{21} = 35$ (Non-smokers with lung cancer)
* $n_{22} = 61$ (Non-smokers without lung cancer)
* $N = n_{11} + n_{12} + n_{21} + n_{22} = 322$ (Total sample size)

## 1. a) [5 points] The proportion with cancer in the population is estimated by (126+35)/(126+35+100+61)=0.5.

False

$\hat{p}_{cancer} = \frac{n_{11} + n_{21}}{N} = \frac{126 + 35}{322} = 0.5$
```{r}
cancer_yes <- 126 + 35
total <- 126 + 35 + 100 + 61
p_cancer <- cancer_yes/total
cat("Proportion with cancer(sampled):", round(p_cancer, 3))
```

The reason this is False is because it is a case-control study where cases and controls were sampled separately. The proportion 0.5 is by design and doesn't estimate the true population proportion. In simpler terms, the provided calculation estimates the proportion of lung cancer in the sample, not the population.  These case-control studies are designed to oversample cases (people with the disease) to have enough data for analysis. So the sample proportions do not reflect the population proportions. Hence one cannot estimate the population prevalence of lung cancer from a case-control study.

## b) [5 points] The proportion of the population that smokes is estimated by (126+100)/ (126+35+100+61)=0.702.

$\hat{p}{smoker} = \frac{n{11} + n_{12}}{N} = \frac{126 + 100}{322} = 0.702$

False

```{r}
smokers <- 126 + 100
p_smokers <- smokers/total
cat("Proportion of smokers(sampled):", round(p_smokers, 3))
```

Similar to 1a previously, this calculation estimates the proportion of smokers in the sample, not the population.  The sampling scheme of a case-control study distorts the proportions of exposures (like smoking) in the sample compared to the population.

## c) [5 points] The probability of having lung cancer among Smokers is estimated by 126/226=0.558.

$\hat{p}{cancer|smoker} = \frac{n{11}}{n_{11} + n_{12}} = \frac{126}{226} = 0.558$

True

```{r}
cancer_smokers <- 126
total_smokers <- 226
cancer_given_smoker <- cancer_smokers / total_smokers
cat("P(cancer|smoker):", round(cancer_given_smoker, 3))
```

So assuming that the this question part does not talk about the population but now talking about this sample case study, so this calculation is correct for estimating the probability of having lung cancer among smokers in the study sample. One can divide the number of smokers with lung cancer (126) by the total number of smokers (126 + 100 = 226).
In layman’s terms, if one looks only at the smokers in the study, approximately 55.8% of them have lung cancer.


## d) [5 points] The probability of having lung cancer among Non-Smokers is estimated by 35/96=0.365.

$\hat{p}{cancer|non-smoker} = \frac{n{21}}{n_{21} + n_{22}} = \frac{35}{96} = 0.365$

True

```{r}
cancer_nonsmokers <- 35
total_nonsmokers <- 96
cancer_given_nonsmoker <- cancer_nonsmokers / total_nonsmokers
cat("P(cancer|non-smoker):", round(cancer_given_nonsmoker, 3))
```

Again, assuming that the this question part does not talk about the population but now talking about this sample case study, this calculation is correct for estimating the probability of having lung cancer among non-smokers in the study sample. One can divide the number of non-smokers with lung cancer (35) by the total number of non-smokers (35 + 61 = 96).
In layman’s terms, if one looks only at the non-smokers in this study, approximately 36.5% of them have lung cancer.

## e) [5 points] The relative risk of having lung cancer, Smokers relative to non-Smokers is 0.558/0.365=1.529.

$RR = \frac{\hat{p}{cancer|smoker}}{\hat{p}{cancer|non-smoker}} = \frac{0.558}{0.365} = 1.529$

False

```{r}
relative_risk <- cancer_given_smoker / cancer_given_nonsmoker
cat("Relative Risk:", round(relative_risk, 3))
```

In a case-control study, one cannot directly calculate the relative risk because the incidence rates in the exposed and unexposed groups are not available. Since, in class theory, I remember that Relative risk is the ratio of the probability of disease in the exposed group to the probability of disease in the unexposed group.  Now simply, because these case-control studies don't give the true population probabilities of having the disease, one can't calculate the relative risk.

## f) [5 points] The odds ratio of having lung cancer for smokers relative to non-smokers is (12661)/(35100)=2.196.

$OR = \frac{n_{11}n_{22}}{n_{12}n_{21}} = \frac{126 \times 61}{100 \times 35} = 2.196$

True
```{r}
OR <- (126*61)/(100*35)
cat("Odds Ratio:", round(OR, 3))
```

Now the Odds ratio is valid for case-control studies as it doesn't require probability estimation. So the correct calculation steps will be :
1) First, calculating odds for smokers:
$Odds_{smoker} = \frac{n_{11}}{n_{12}} = \frac{126}{100} = 1.26$

2) Calculating odds for non-smokers:
$Odds_{non-smoker} = \frac{n_{21}}{n_{22}} = \frac{35}{61} = 0.574$

3) The odds ratio is:
$OR = \frac{Odds_{smoker}}{Odds_{non-smoker}} = \frac{n_{11}n_{22}}{n_{12}n_{21}} = \frac{126 \times 61}{100 \times 35} = 2.196$



## g) [10 points] Now you must find the 95% CI for the odds ratio from these data.