---
title: "Homework 4: Namit Shrivastava"
format: pdf
editor: visual
---

## Problem Statement

Your client is a retail banking institution. Term deposits are a major source of income for a bank. A term deposit is a cash investment held at a financial institution. Your money is invested for an agreed rate of interest over a fixed amount of time, or term. 

The bank has various outreach plans to sell term deposits to their customers such as email marketing, advertisements, telephonic marketing and digital marketing. Telephonic marketing campaigns still remain one of the most effective way to reach out to people. However, they require huge investment as large call centers are hired to actually execute these campaigns. 

Hence, it is crucial to identify the customers most likely to convert beforehand so that they can be specifically targeted via call. You are provided with the client data such as: age of the client, their job type, their marital status, etc. Along with the client data, you are also provided with the information of the call such as the duration of the call, day and month of the call, etc. Given this information, your task is to predict if the client will subscribe to term deposit.

## Data Dictionary
$$

```{=tex}
\begin{center}
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Variable} & \textbf{Definition} \\
\hline
ID & Unique client ID \\
\hline
age & Age of the client \\
\hline
job & Type of job \\
\hline
marital & Marital status of the client \\
\hline
education & Education level \\
\hline
default & Credit in default \\
\hline
balance & Average yearly balance \\
\hline
housing & Housing loan \\
\hline
loan & Personal loan \\
\hline
contact & Type of communication \\
\hline
month & Contact month \\
\hline
day & Day of week of contact \\
\hline
duration & Contact duration \\
\hline
campaign & Number of contacts performed during this campaign to the client \\
\hline
pdays & Number of days that passed by after the client was last contacted \\
\hline
previous & Number of contacts performed before this campaign \\
\hline
poutcome & Outcome of the previous marketing campaign \\
\hline
subscribed & Has the client subscribed a term deposit? (target variable) \\
\hline
\end{tabular}
\end{center}
$$

## 1a. [10 points] Create an empirical logit plot with the response variable (y) subscribed by the predictor variable (x) age. You may want to convert subscribed to a numeric variable first.

Let the proportion of successes be:
$$
p = \frac{\text{Number of Successes}}{\text{Total Observations}}
$$

Then we add a continuity correction term of \( \frac{1}{2n} \) in the numerator and denominator:

$$
p^* = p + \frac{1}{2n}, \quad 1 - p^* = 1 - p + \frac{1}{2n}
$$

Thus the empirical logit is defined as:
$$
\text{empirical logit} 
= \ln \left( \frac{p^*}{1 - p^*} \right)
= \ln \left( \frac{p + \frac{1}{2n}}{1 - p + \frac{1}{2n}} \right)
$$

```{r}
# Loading libraries
library(dplyr)
library(ggplot2)

# Reading data
deposit_data <- read.csv("deposit-1.csv")

# Converting subscribed to numeric
deposit_data$subscribed_num <- ifelse(deposit_data$subscribed == "yes", 1, 0)

# Calculating empirical logits by age
empirical_logits <- deposit_data %>%
  group_by(age) %>%
  summarize(
    n = n(),
    successes = sum(subscribed_num),
    p = successes / n,
    emp_logit = log((p + 1/(2*n)) / (1 - p + 1/(2*n)))
  )

# Creating the empirical logit plot
ggplot(empirical_logits, aes(x = age, y = emp_logit)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Empirical Logit Plot",
    x = "Age",
    y = "Empirical Logit"
  ) +
  theme_minimal()
```

## 1b. [10 points] Report the proportion (tabular format is fine) subscribed=”yes” for each of the categories for job, marital, education, default, housing, loan, and contact.

So, the proportion of subscribed="yes" for each category is calculated as:

$$
\text{Proportion} = \frac{\text{Number of subscribed="yes"}}{\text{Total observations in category}}
$$


```{r}
library(knitr)
# Function to calculate proportions for a categorical variable
calculate_proportions <- function(data, var_name) {
  props <- data %>%
    group_by(!!sym(var_name)) %>%
    summarize(
      total = n(),
      subscribed_yes = sum(subscribed == "yes"),
      proportion = round(subscribed_yes / total, 4)
    ) %>%
    arrange(desc(proportion)) %>%
    mutate(variable = var_name)

  return(props)
}

# List of categorical variables
cat_vars <- c("job", "marital", "education", "default", "housing", "loan", "contact")

# Calculating proportions for each variable
proportion_tables <- lapply(cat_vars, function(var) {
  calculate_proportions(deposit_data, var)
})

# Printing tables with knitr for each variable
for(i in seq_along(cat_vars)) {
  cat("\n### Proportions for", cat_vars[i], "\n")
  print(
    kable(
      proportion_tables[[i]],
      col.names = c("Category", "Total", "Subscribed Yes", "Proportion", "Variable"),
      caption = paste("Subscription proportions by", cat_vars[i]),
      digits = 4
    )
  )
  cat("\n")
}
```

Now, the tables above show the proportion of customers who subscribed to a term deposit ("yes") for each category within the specified variables. Key observations are

1. By job type: Which occupations show higher subscription rates
2. By marital status: How relationship status affects subscription likelihood
3. By education level: The impact of education on subscription rates
4. By default, housing, and loan status: How existing financial commitments relate to subscription rates
5. By contact type: Which communication method yields better subscription rates

Based on my analysis of the subscription proportions across different categories, the table observations are:

1. For job types, students and retirees have the highest subscription rates at 28.7% and 23% respectively, which makes sense as they might be more focused on saving and investment. On the other hand, blue-collar workers and entrepreneurs have the lowest subscription rates, around 7-8%.

2. When looking at marital status, single individuals are more likely to subscribe (15.1%) compared to married (10%) or divorced (12.3%) individuals. This could be because single people might have fewer financial commitments and more disposable income.

3. Education level also plays a significant role. Those with tertiary education show the highest subscription rate (15.2%), while those with primary education have the lowest (8.9%). This suggests that higher education levels might be associated with better financial literacy and a greater propensity to invest.

4. Financial circumstances are crucial too. Customers without existing financial commitments are more likely to subscribe. Those without housing loans (16.8%) or personal loans (12.8%) show higher subscription rates than those with such commitments. Similarly, customers with no credit defaults have a much higher subscription rate (11.8%) compared to those with defaults (7%).

5. Regarding contact methods, cellular phone contact proves most effective with a 15% success rate, compared to telephone (13%) and unknown contact methods (4.1%). This suggests that mobile phone communication might be the most effective channel for future campaigns.


## 1c. [10 points] Plot the response variable, proportion subscribed=”yes”, for each of the following values of campaign: 1, 2, 3, 4, 5, 6+.

```{r}
# Creating campaign groups
deposit_data <- deposit_data %>%
  mutate(campaign_group = case_when(
    campaign >= 6 ~ "6+",
    TRUE ~ as.character(campaign)
  ))

# Calculating proportions for each campaign group
campaign_props <- deposit_data %>%
  group_by(campaign_group) %>%
  summarize(
    total = n(),
    subscribed_yes = sum(subscribed == "yes"),
    proportion = round(subscribed_yes / total, 4)
  ) %>%
  mutate(campaign_group = factor(campaign_group, 
                                 levels = c("1", "2", "3", "4", "5", "6+")))

# Creating the bar plot of proportions
library(ggplot2)
ggplot(campaign_props, aes(x = campaign_group, y = proportion)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)),
            vjust = -0.5) +
  labs(
    title = "Proportion of Subscriptions by Number of Campaign Contacts",
    x = "Number of Campaign Contacts",
    y = "Proportion Subscribed",
    caption = "Note: 6+ represents six or more contacts"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)
```

From the bar plot:

1. The proportion of clients subscribing decreases as the number of campaign contacts increases. For example:

* The highest subscription rate is observed at 1 contact.

* The subscription rate continues to drop with more contacts, reaching its lowest for 6+ contacts.

2. Now the diminishing returns with additional contacts suggests that beyond a certain point, increasing the number of calls may not be effective in converting clients and could even have a negative impact.

3. Focusing on fewer but more targeted contacts might be a more efficient strategy for maximizing subscriptions while minimizing costs.


## 1d. [15 points] Next, we want to evaluate if campaign contacts are effective. Estimate a logistic regression model using the variable campaign as a predictor of subscribed=yes. Are more campaign contacts effective at producing subscriptions to term deposits?

So to evaluate if campaign contacts are effective at producing subscriptions, let me fit a logistic regression model:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \times \text{campaign}
$$

where $p$ is the probability of subscribing to a term deposit.

To evaluate if campaign contacts are effective at producing subscriptions, let's fit a logistic regression model:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \times \text{campaign}
$$

where $p$ is the probability of subscribing to a term deposit.

```{r}
deposit_data$subscribed_num <- ifelse(deposit_data$subscribed == "yes", 1, 0)

# Fitting a logistic regression model with 'campaign' as the only predictor
logistic_model <- glm(subscribed_num ~ campaign,
                      data = deposit_data,
                      family = binomial)

# Summarizing the model
model_summary <- summary(logistic_model)
print(model_summary)

# Calculating odds ratios and confidence intervals
odds_ratios <- exp(coef(logistic_model))
conf_int <- exp(confint(logistic_model))

# Combining results in a data frame
model_results <- data.frame(
  Coefficient = coef(logistic_model),
  Std_Error = model_summary$coefficients[, "Std. Error"],
  z_value = model_summary$coefficients[, "z value"],
  p_value = model_summary$coefficients[, "Pr(>|z|)"],
  Odds_Ratio = odds_ratios,
  CI_Lower = conf_int[, 1],
  CI_Upper = conf_int[, 2]
)
print(model_results)
```

Based on the logistic regression model results, I can see that the coefficient for the campaign variable is -0.1225, which is statistically significant with a p-value less than 0.0001. This negative coefficient indicates that each additional campaign contact is associated with a decrease in the log-odds of a customer subscribing to a term deposit.

The odds ratio for the campaign variable is approximately 0.885. This means that for each additional campaign contact, the odds of a customer subscribing to a term deposit decrease by about 11.5% (since 1 - 0.885 = 0.115). The 95% confidence interval for the odds ratio ranges from 0.868 to 0.901, which does not include 1, further confirming the statistical significance of this finding.

Ok so in practical terms, this suggests that increasing the number of campaign contacts is not effective in producing more subscriptions. In fact, it appears to have a diminishing return, where more contacts actually reduce the likelihood of a customer subscribing. Interestingly, this insight can help in optimizing marketing strategies by focusing on fewer, more impactful contacts rather than repeatedly contacting the same customers.


## 1e [5 points] What is the probability of a person with zero contacts (i.e. campaign=0) subscribing to a term deposit? What is the probability of a person with one contacts (i.e. campaign=1) subscribing to a term deposit? What is the probability of a person with two contacts (i.e. campaign=2) subscribing to a term deposit?


Using the fitted logistic regression model, one can calculate the probability of subscription for different numbers of contacts using the formula:

$$
P(\text{subscription}) = \frac{e^{\beta_0 + \beta_1 \times \text{campaign}}}{1 + e^{\beta_0 + \beta_1 \times \text{campaign}}}
$$

```{r}
deposit_data$subscribed_num <- ifelse(deposit_data$subscribed == "yes", 1, 0)

# Creating new data for prediction
new_data <- data.frame(campaign = c(0, 1, 2))

# Computing predicted probabilities
predicted_probs <- predict(logistic_model,
                           newdata = new_data,
                           type = "response")

# Combining results into a table
prob_table <- data.frame(
  Contacts = new_data$campaign,
  Probability = predicted_probs
)

print(prob_table)
```

Based on the logistic regression model, I calculated the probabilities of a person subscribing to a term deposit given different numbers of campaign contacts. For a person with zero contacts, the probability of subscribing is approximately 15.2%. If the person is contacted once, the probability decreases to about 13.7%. With two contacts, the probability further decreases to around 12.3%.

This result aligns with the earlier findings that increasing the number of campaign contacts tends to reduce the likelihood of a customer subscribing to a term deposit. It suggests that the first contact is the most impactful, and additional contacts may not only fail to increase the subscription rate but might actually deter potential subscribers. This insight can help in refining marketing strategies to focus on making the initial contact as effective as possible, rather than relying on multiple follow-up calls.


## 1f. [20 points] Estimate a logistic regression model using the variable campaign as a predictor along with the following other variables: job, marital, education, default, housing, loan, contact, age, and campaign. Consider the form in which age should enter the model (i.e. categorical, continuous, transformed) and choose the best option for this model. Are more campaign contacts effective at producing subscriptions to term deposits conditional on the additional predictors?


First, let me examine age distribution to determine its appropriate form in the model:

```{r}
library(dplyr)
library(ggplot2)
library(car)

# Reading data
deposit <- read.csv("deposit-1.csv")
# Converting subscribed to numeric
deposit$subscribed_numeric <- ifelse(deposit$subscribed == "yes", 1, 0)

# Plotting a histogram to check the distribution of age
ggplot(deposit, aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Age",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()

# Converting categorical variables to factors
deposit <- deposit %>%
  mutate(
    job = as.factor(job),
    marital = as.factor(marital),
    education = as.factor(education),
    default = as.factor(default),
    housing = as.factor(housing),
    loan = as.factor(loan),
    contact = as.factor(contact)
  )

# Fitting the logistic regression model with all relevant predictors
model_full <- glm(
  subscribed_numeric ~ campaign + job + marital + education + default + 
    housing + loan + contact + age,
  data = deposit,
  family = binomial
)

# Summarizing the model
summary(model_full)

# Checking for multicollinearity
vif(model_full)
```
From this model, each additional campaign contact corresponds to a decrease in the odds of subscribing (negative coefficient). 

Meanwhile, certain job categories, like retirees and students, show higher odds of subscription, whereas blue-collar workers and housemaids exhibit lower odds. 
Married individuals are slightly less likely to subscribe than singles, and having housing or personal loans further reduces the likelihood of subscribing. 

Age has a small but positive effect, suggesting that older clients have a marginally higher chance of subscribing. Overall, the model seems to fit well, and the VIF values indicate no serious multicollinearity concerns.


## 1g. [10 points] For the model estimated in 1f, what is an interpretation of the coefficient for campaign?

## Interpretation of Campaign Coefficient

The coefficient for campaign in the multivariate logistic model represents the change in log-odds of subscription for each additional contact, holding all other variables constant.

```{r}
# Extracting campaign coefficient and related statistics
campaign_coef <- coef(full_model)["campaign"]
campaign_OR <- exp(campaign_coef)
campaign_CI <- exp(confint(full_model))["campaign", ]

# Creating detailed interpretation table
interpretation_stats <- data.frame(
  Metric = c("Log-odds coefficient (β)", 
             "Odds Ratio", 
             "Percentage change in odds",
             "95% CI Lower",
             "95% CI Upper",
             "p-value"),
  Value = c(
    round(campaign_coef, 4),
    round(campaign_OR, 4),
    round((campaign_OR - 1) * 100, 2),
    round(campaign_CI[1], 4),
    round(campaign_CI[2], 4),
    format.pval(summary(full_model)$coefficients["campaign", "Pr(>|z|)"], digits = 4)
  )
)

# Displaying interpretation statistics
knitr::kable(
  interpretation_stats,
  caption = "Campaign Coefficient Interpretation Statistics",
  align = c('l', 'r')
)
```

Now talking about the mathematical interpretation

The logistic regression equation for the campaign variable, holding other variables constant:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_{\text{campaign}} \times \text{campaign} + \text{other variables}
$$

where $\beta_{\text{campaign}} = `r round(campaign_coef, 4)`$

Now coming to the practical interpretations:

1. **Log-odds scale**: 
   - For each additional campaign contact, the log-odds of subscription decrease by `r abs(round(campaign_coef, 4))` units, holding other variables constant

2. **Odds scale**: 
   - The odds ratio of `r round(campaign_OR, 4)` means that for each additional campaign contact, the odds of subscription decrease by `r abs(round((1-campaign_OR)*100, 1))`%, controlling for other variables
   
3. **Statistical significance**:
   - The effect is statistically significant (p < `r format.pval(summary(full_model)$coefficients["campaign", "Pr(>|z|)"], digits=4)`)
   - We can be 95% confident that the true odds ratio lies between `r round(campaign_CI[1], 4)` and `r round(campaign_CI[2], 4)`

So in a business perspective, even after accounting for demographic and financial characteristics:
- Additional campaign contacts are associated with lower subscription probability
- Each contact reduces the odds of subscription by approximately `r abs(round((1-campaign_OR)*100, 1))`%
- The negative relationship is consistent and statistically significant

## 1h. [10 points] Use the likelihood ratio test discussed in class to evaluate whether the model in 1f is a better fit than the model in 1d.

## 1h. Likelihood Ratio Test

Now I will compare the simpler model (1d) with campaign only to the full model (1f) with multiple predictors using the likelihood ratio test:

$$
LR = -2(\ln L_{\text{simple}} - \ln L_{\text{full}}) \sim \chi^2_{df}
$$

where $df$ is the difference in degrees of freedom between the models.

```{r}
# Performing likelihood ratio test
lr_test <- anova(logistic_model, full_model, test = "LRT")

# Creating detailed results table
lr_results <- data.frame(
  Model = c("Simple (1d)", "Full (1f)"),
  DF = c(lr_test$Df[1], lr_test$Df[2]),
  Deviance = c(lr_test$Deviance[1], lr_test$Deviance[2]),
  AIC = c(AIC(logistic_model), AIC(full_model)),
  BIC = c(BIC(logistic_model), BIC(full_model))
)

# Displaying model comparison statistics
knitr::kable(
  lr_results,
  caption = "Model Comparison Statistics",
  digits = 2,
  align = c('l', 'r', 'r', 'r', 'r')
)

# Creating likelihood ratio test results table
lr_stats <- data.frame(
  Metric = c("Likelihood Ratio Statistic", 
             "Degrees of Freedom", 
             "p-value",
             "Change in Deviance"),
  Value = c(
    lr_test$`LR stat`[2],
    lr_test$`Df`[2] - lr_test$`Df`[1],
    lr_test$`Pr(>Chi)`[2],
    lr_test$Deviance[1] - lr_test$Deviance[2]
  )
)

# Displaying likelihood ratio test results
knitr::kable(
  lr_stats,
  caption = "Likelihood Ratio Test Results",
  digits = 4,
  align = c('l', 'r')
)
```

Interpretting the results:

1. **Test Statistic**:
   - LR = `r round(lr_test$"LR stat"[2], 2)`
   - Degrees of freedom = `r lr_test$"Df"[2] - lr_test$"Df"[1]`
   - p-value < `r format.pval(lr_test$"Pr(>Chi)"[2], digits=4)`

2. **Model Comparison**:
   - Reduction in deviance: `r round(lr_test$Deviance[1] - lr_test$Deviance[2], 2)`
   - The full model has lower AIC (`r round(AIC(full_model), 2)` vs `r round(AIC(logistic_model), 2)`)
   - The full model has lower BIC (`r round(BIC(full_model), 2)` vs `r round(BIC(logistic_model), 2)`)

3. **Conclusion**: 
   The likelihood ratio test shows strong evidence (p < `r format.pval(lr_test$"Pr(>Chi)"[2], digits=4)`) that the full model provides a significantly better fit than the simple model. The substantial reduction in deviance and lower information criteria (AIC, BIC) further support using the more complex model for predicting term deposit subscriptions.

