---
title: "Homework 4: Namit Shrivastava"
format: pdf
editor: visual
---

## Problem Statement

Your client is a retail banking institution. Term deposits are a major source of income for a bank. A term deposit is a cash investment held at a financial institution. Your money is invested for an agreed rate of interest over a fixed amount of time, or term. 

The bank has various outreach plans to sell term deposits to their customers such as email marketing, advertisements, telephonic marketing and digital marketing. Telephonic marketing campaigns still remain one of the most effective way to reach out to people. However, they require huge investment as large call centers are hired to actually execute these campaigns. 

Hence, it is crucial to identify the customers most likely to convert beforehand so that they can be specifically targeted via call. You are provided with the client data such as: age of the client, their job type, their marital status, etc. Along with the client data, you are also provided with the information of the call such as the duration of the call, day and month of the call, etc. Given this information, your task is to predict if the client will subscribe to term deposit.

## Data Dictionary
$$

```{=tex}
\begin{center}
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Variable} & \textbf{Definition} \\
\hline
ID & Unique client ID \\
\hline
age & Age of the client \\
\hline
job & Type of job \\
\hline
marital & Marital status of the client \\
\hline
education & Education level \\
\hline
default & Credit in default \\
\hline
balance & Average yearly balance \\
\hline
housing & Housing loan \\
\hline
loan & Personal loan \\
\hline
contact & Type of communication \\
\hline
month & Contact month \\
\hline
day & Day of week of contact \\
\hline
duration & Contact duration \\
\hline
campaign & Number of contacts performed during this campaign to the client \\
\hline
pdays & Number of days that passed by after the client was last contacted \\
\hline
previous & Number of contacts performed before this campaign \\
\hline
poutcome & Outcome of the previous marketing campaign \\
\hline
subscribed & Has the client subscribed a term deposit? (target variable) \\
\hline
\end{tabular}
\end{center}
$$

## 1a. [10 points] Create an empirical logit plot with the response variable (y) subscribed by the predictor variable (x) age. You may want to convert subscribed to a numeric variable first.

The empirical logit is calculated using the formula:

$$
\text{empirical logit} = \ln\left(\frac{p + \frac{1}{2n}}{1 - p + \frac{1}{2n}}\right)
$$

where:
- $p$ is the proportion of successes in each group
- $n$ is the number of observations in each group
```{r}
# Loading the required libraries
library(dplyr)
library(ggplot2)

# Reading the data
deposit_data <- read.csv("deposit-1.csv")

# Converting subscribed to numeric
deposit_data$subscribed_num <- ifelse(deposit_data$subscribed == "yes", 1, 0)

# Calculating empirical logits by age
empirical_logits <- deposit_data %>%
  group_by(age) %>%
  summarize(
    n = n(),
    successes = sum(subscribed_num),
    p = successes/n,
    emp_logit = log((p + 1/(2*n))/(1 - p + 1/(2*n)))
  )

# Creating the empirical logit plot
ggplot(empirical_logits, aes(x = age, y = emp_logit)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Empirical Logit Plot",
    x = "Age",
    y = "Empirical Logit",
    caption = "Note: Blue line shows LOESS smooth with 95% confidence interval"
  ) +
  theme_minimal()

# Displaying first few rows of calculations
knitr::kable(
  head(empirical_logits),
  caption = "Sample of Empirical Logit Calculations",
  digits = 4
)
```

## 1b. [10 points] Report the proportion (tabular format is fine) subscribed=”yes” for each of the categories for job, marital, education, default, housing, loan, and contact.

So, the proportion of subscribed="yes" for each category is calculated as:

$$
\text{Proportion} = \frac{\text{Number of subscribed="yes"}}{\text{Total observations in category}}
$$


```{r}

# Function to calculate proportions for a categorical variable
calculate_proportions <- function(data, var_name) {
  props <- data %>%
    group_by(!!sym(var_name)) %>%
    summarize(
      total = n(),
      subscribed_yes = sum(subscribed == "yes"),
      proportion = round(subscribed_yes/total, 4)
    ) %>%
    arrange(desc(proportion))
  
  # Add variable name to the results
  props <- props %>%
    mutate(variable = var_name) %>%
    select(variable, everything())
  
  return(props)
}

# List of categorical variables
cat_vars <- c("job", "marital", "education", "default", "housing", "loan", "contact")

# Calculating proportions for each variable
proportion_tables <- lapply(cat_vars, function(var) {
  calculate_proportions(deposit_data, var)
})

# Displaying tables with proper formatting
for(i in seq_along(cat_vars)) {
  cat("\n### Proportions for", cat_vars[i], "\n")
  print(knitr::kable(
    proportion_tables[[i]], 
    col.names = c("Variable", "Category", "Total", "Subscribed Yes", "Proportion"),
    caption = paste("Subscription proportions by", cat_vars[i]),
    digits = 4
  ))
  cat("\n")
}
```

Now, the tables above show the proportion of customers who subscribed to a term deposit ("yes") for each category within the specified variables. Key observations are

1. By job type: Which occupations show higher subscription rates
2. By marital status: How relationship status affects subscription likelihood
3. By education level: The impact of education on subscription rates
4. By default, housing, and loan status: How existing financial commitments relate to subscription rates
5. By contact type: Which communication method yields better subscription rates

Based on my analysis of the subscription proportions across different categories, the table observations are:

1. Students and retirees are most likely to subscribe to term deposits, with subscription rates of 28.7% and 23% respectively. This might be because they're more focused on saving and investment. In contrast, blue-collar workers and entrepreneurs show the lowest subscription rates (around 7-8%).

2. Looking at marital status, single individuals are more likely to subscribe (15.1%) compared to married (10%) or divorced (12.3%) individuals. Education level also plays a significant role - those with tertiary education show the highest subscription rate (15.2%), while those with primary education have the lowest (8.9%).

3. Financial circumstances appear crucial: customers without existing financial commitments are more likely to subscribe. Those without housing loans (16.8%) or personal loans (12.8%) show higher subscription rates than those with such commitments. Similarly, customers with no credit defaults have a much higher subscription rate (11.8%) compared to those with defaults (7%).

4. Regarding contact methods, cellular phone contact proves most effective with a 15% success rate, compared to telephone (13%) and unknown contact methods (4.1%). This suggests that mobile phone communication might be the most effective channel for future campaigns.


## 1c. [10 points] Plot the response variable, proportion subscribed=”yes”, for each of the following values of campaign: 1, 2, 3, 4, 5, 6+.

```{r}
# Creating campaign groups
deposit_data <- deposit_data %>%
  mutate(campaign_group = case_when(
    campaign >= 6 ~ "6+",
    TRUE ~ as.character(campaign)
  ))

# Calculating proportions for each campaign group
campaign_props <- deposit_data %>%
  group_by(campaign_group) %>%
  summarize(
    total = n(),
    subscribed_yes = sum(subscribed == "yes"),
    proportion = round(subscribed_yes/total, 4)
  ) %>%
  mutate(campaign_group = factor(campaign_group, 
                                levels = c("1", "2", "3", "4", "5", "6+")))

# Creating the plot
ggplot(campaign_props, aes(x = campaign_group, y = proportion)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)),
            vjust = -0.5) +
  labs(
    title = "Proportion of Subscriptions by Number of Campaign Contacts",
    x = "Number of Contacts",
    y = "Proportion Subscribed",
    caption = "Note: 6+ represents six or more contacts"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent, limits = c(0, max(campaign_props$proportion) * 1.2))

# Displaying the summary table
knitr::kable(
  campaign_props,
  col.names = c("Campaign Contacts", "Total Customers", "Subscribed", "Proportion"),
  caption = "Subscription rates by number of campaign contacts",
  digits = 4
)
```



Alright, so looking at the results of the campaign contact analysis, the most striking finding is that our first contact attempt is actually our most successful, with about 14.5% of customers subscribing. There's a clear downward trend after that the success rate drops with each additional contact. By the second contact, it is down to 11.4%, and it keeps declining steadily.

What's particularly concerning is that when I reach out 6 or more times, the success rate plummets to just 6.2%. This suggests that persistently contacting reluctant customers might not be the best strategy. In fact, one might be better off focusing these resources on making that first contact count, rather than making repeated attempts with the same customers.

This data tells me one should probably rethink the follow-up strategy. Instead of making multiple attempts with the same customers, one might get better results by:

1. Investing more in making that first contact as effective as possible
2. Moving on to fresh prospects after 2-3 unsuccessful attempts
3. Analyzing what makes the first contacts more successful and applying those lessons

## 1d. [15 points] Next, we want to evaluate if campaign contacts are effective. Estimate a logistic regression model using the variable campaign as a predictor of subscribed=yes. Are more campaign contacts effective at producing subscriptions to term deposits?

So to evaluate if campaign contacts are effective at producing subscriptions, let me fit a logistic regression model:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \times \text{campaign}
$$

where $p$ is the probability of subscribing to a term deposit.

To evaluate if campaign contacts are effective at producing subscriptions, let's fit a logistic regression model:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \times \text{campaign}
$$

where $p$ is the probability of subscribing to a term deposit.

```{r}
# Fitting logistic regression model
logistic_model <- glm(subscribed_num ~ campaign, 
                      data = deposit_data,
                      family = binomial)

# Creating summary statistics
summary_stats <- summary(logistic_model)
odds_ratios <- exp(coef(logistic_model))
conf_intervals <- exp(confint(logistic_model))

# Combining results into a table
model_results <- data.frame(
  Coefficient = coef(logistic_model),
  Std_Error = summary_stats$coefficients[,2],
  z_value = summary_stats$coefficients[,3],
  p_value = summary_stats$coefficients[,4],
  Odds_Ratio = odds_ratios,
  CI_Lower = conf_intervals[,1],
  CI_Upper = conf_intervals[,2]
)

# Model fit statistics
model_fit <- data.frame(
  Metric = c("Null Deviance", "Residual Deviance", "AIC", "McFadden R²"),
  Value = c(
    logistic_model$null.deviance,
    logistic_model$deviance,
    logistic_model$aic,
    1 - (logistic_model$deviance/logistic_model$null.deviance)
  )
)

# Visualizing predicted probabilities
campaign_range <- data.frame(campaign = 1:10)
predicted_probs <- predict(logistic_model, 
                          newdata = campaign_range, 
                          type = "response")

pred_df <- data.frame(
  campaign = campaign_range$campaign,
  prob = predicted_probs
)

# Creating visualization
ggplot(pred_df, aes(x = campaign, y = prob)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Predicted Probability of Subscription by Number of Campaigns",
    x = "Number of Campaign Contacts",
    y = "Predicted Probability of Subscription"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)

# Display results
print(model_results)
print(model_fit)
```

Now interpretting the results obtained, 

1. **Model Coefficients**:
   - The coefficient for campaign (B1) is `r round(coef(logistic_model)[2], 4)`, indicating a negative relationship
   - This translates to an odds ratio of `r round(odds_ratios[2], 4)`, meaning the odds of subscription decrease by `r round((1-odds_ratios[2])*100, 1)`% for each additional contact

2. **Statistical Significance**:
   - The p-value for campaign (`r format.pval(summary_stats$coefficients[2,4], digits=4)`) indicates the relationship is statistically significant
   - The 95% confidence interval for the odds ratio doesn't include 1, confirming the significance

3. **Model Fit**:
   - McFadden's $R^2$ of `r round(1 - (logistic_model$deviance/logistic_model$null.deviance), 4)` suggests moderate predictive power
   - The negative coefficient and visualization demonstrate that more campaign contacts are associated with lower subscription probability

So by concluding, the analysis strongly suggests that increasing the number of campaign contacts is not effective at producing subscriptions. In fact, each additional contact is associated with significantly lower odds of subscription, supporting our earlier findings about the diminishing returns of repeated contacts.


## 1e [5 points] What is the probability of a person with zero contacts (i.e. campaign=0) subscribing to a term deposit? What is the probability of a person with one contacts (i.e. campaign=1) subscribing to a term deposit? What is the probability of a person with two contacts (i.e. campaign=2) subscribing to a term deposit?

## 1e. Probability Calculations for Different Contact Numbers

Using the fitted logistic regression model, one can calculate the probability of subscription for different numbers of contacts using the formula:

$$
P(\text{subscription}) = \frac{e^{\beta_0 + \beta_1 \times \text{campaign}}}{1 + e^{\beta_0 + \beta_1 \times \text{campaign}}}
$$

```{r}
# Creating new data for predictions
new_data <- data.frame(campaign = c(0, 1, 2))

# Calculating predicted probabilities
predicted_probs <- predict(logistic_model, 
                          newdata = new_data, 
                          type = "response")

# Combining results
prob_table <- data.frame(
  Contacts = new_data$campaign,
  Probability = predicted_probs
)

# Displaying results in formatted table
knitr::kable(
  prob_table,
  caption = "Predicted Probabilities of Subscription by Number of Contacts",
  col.names = c("Number of Contacts", "Probability of Subscription"),
  digits = 4,
  align = c('c', 'c')
)

# Calculating log-odds for each scenario
log_odds <- predict(logistic_model, newdata = new_data, type = "link")
calculations <- data.frame(
  Contacts = new_data$campaign,
  Log_Odds = log_odds,
  Odds = exp(log_odds),
  Probability = predicted_probs
)

# Displaying detailed calculations
knitr::kable(
  calculations,
  caption = "Step-by-step Probability Calculations",
  col.names = c("Contacts", "Log-Odds", "Odds", "Probability"),
  digits = 4,
  align = c('c', 'c', 'c', 'c')
)
```


## 1f. [20 points] Estimate a logistic regression model using the variable campaign as a predictor along with the following other variables: job, marital, education, default, housing, loan, contact, age, and campaign. Consider the form in which age should enter the model (i.e. categorical, continuous, transformed) and choose the best option for this model. Are more campaign contacts effective at producing subscriptions to term deposits conditional on the additional predictors?

## 1f. Multivariate Logistic Regression Analysis

First, let me examine age distribution to determine its appropriate form in the model:

```{r}
# Examining age relationship with empirical logit
ggplot(empirical_logits, aes(x = age, y = emp_logit)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Age vs Empirical Logit",
       x = "Age",
       y = "Empirical Logit") +
  theme_minimal()

# Creating age categories based on quartiles
deposit_data <- deposit_data %>%
  mutate(age_cat = cut(age, 
                       breaks = quantile(age, probs = c(0, 0.25, 0.5, 0.75, 1)),
                       labels = c("Q1", "Q2", "Q3", "Q4"),
                       include.lowest = TRUE))

# Fitting the multivariate logistic regression
full_model <- glm(subscribed_num ~ 
                  age_cat +          # Age as categorical (based on quartiles)
                  job + 
                  marital + 
                  education + 
                  default + 
                  housing + 
                  loan + 
                  contact + 
                  campaign,
                data = deposit_data,
                family = binomial)

# Creating summary with odds ratios
summary_stats <- summary(full_model)
odds_ratios <- exp(coef(full_model))
conf_intervals <- exp(confint(full_model))

# Combining results
model_results <- data.frame(
  Variable = names(coef(full_model)),
  Coefficient = coef(full_model),
  Std_Error = summary_stats$coefficients[,2],
  z_value = summary_stats$coefficients[,3],
  p_value = summary_stats$coefficients[,4],
  Odds_Ratio = odds_ratios,
  CI_Lower = conf_intervals[,1],
  CI_Upper = conf_intervals[,2]
)

# Displaying results
knitr::kable(
  model_results,
  caption = "Multivariate Logistic Regression Results",
  digits = 4,
  align = 'c',
  col.names = c("Variable", "β", "SE", "z-value", "p-value", "OR", "95% CI Lower", "95% CI Upper")
)

# Model fit statistics
model_fit <- data.frame(
  Metric = c("Null Deviance", "Residual Deviance", "AIC", "McFadden R²"),
  Value = c(
    full_model$null.deviance,
    full_model$deviance,
    full_model$aic,
    1 - (full_model$deviance/full_model$null.deviance)
  )
)

knitr::kable(
  model_fit,
  caption = "Model Fit Statistics",
  digits = 4
)

# Visualizing campaign effect while controlling for other variables
pred_data <- expand.grid(
  campaign = 1:10,
  age_cat = "Q2",                # Reference level
  job = "management",            # Most common category
  marital = "married",
  education = "secondary",
  default = "no",
  housing = "no",
  loan = "no",
  contact = "cellular"
)

predicted_probs <- predict(full_model, 
                          newdata = pred_data, 
                          type = "response")

ggplot(data.frame(campaign = pred_data$campaign, 
                  probability = predicted_probs), 
       aes(x = campaign, y = probability)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Adjusted Probability of Subscription by Campaign Contacts",
    x = "Number of Campaign Contacts",
    y = "Predicted Probability of Subscription",
    caption = "Adjusted for other predictors held at reference levels"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)
```

Now interpreting the results obtained:

1. **Age Treatment**:
   - Based on the empirical logit plot, age shows a non-linear relationship
   - Categorized age into quartiles to capture non-linear effects
   
2. **Campaign Effect**:
   - Coefficient for campaign: $\beta_{campaign}$ = `r round(coef(full_model)["campaign"], 4)`
   - Odds ratio = `r round(exp(coef(full_model)["campaign"]), 4)`
   - Still shows negative effect even after controlling for other variables

3. **Model Fit**:
   - McFadden's $R^2$ = `r round(1 - (full_model$deviance/full_model$null.deviance), 4)`
   - Significant improvement over campaign-only model

4. **Key Findings**:
   - Campaign contacts remain negatively associated with subscription probability
   - Effect persists after controlling for demographic and financial characteristics
   - Other significant predictors include age, education, and contact method


## 1g. [10 points] For the model estimated in 1f, what is an interpretation of the coefficient for campaign?

## Interpretation of Campaign Coefficient

The coefficient for campaign in the multivariate logistic model represents the change in log-odds of subscription for each additional contact, holding all other variables constant.

```{r}
# Extracting campaign coefficient and related statistics
campaign_coef <- coef(full_model)["campaign"]
campaign_OR <- exp(campaign_coef)
campaign_CI <- exp(confint(full_model))["campaign", ]

# Creating detailed interpretation table
interpretation_stats <- data.frame(
  Metric = c("Log-odds coefficient (β)", 
             "Odds Ratio", 
             "Percentage change in odds",
             "95% CI Lower",
             "95% CI Upper",
             "p-value"),
  Value = c(
    round(campaign_coef, 4),
    round(campaign_OR, 4),
    round((campaign_OR - 1) * 100, 2),
    round(campaign_CI[1], 4),
    round(campaign_CI[2], 4),
    format.pval(summary(full_model)$coefficients["campaign", "Pr(>|z|)"], digits = 4)
  )
)

# Displaying interpretation statistics
knitr::kable(
  interpretation_stats,
  caption = "Campaign Coefficient Interpretation Statistics",
  align = c('l', 'r')
)
```

Now talking about the mathematical interpretation

The logistic regression equation for the campaign variable, holding other variables constant:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_{\text{campaign}} \times \text{campaign} + \text{other variables}
$$

where $\beta_{\text{campaign}} = `r round(campaign_coef, 4)`$

Now coming to the practical interpretations:

1. **Log-odds scale**: 
   - For each additional campaign contact, the log-odds of subscription decrease by `r abs(round(campaign_coef, 4))` units, holding other variables constant

2. **Odds scale**: 
   - The odds ratio of `r round(campaign_OR, 4)` means that for each additional campaign contact, the odds of subscription decrease by `r abs(round((1-campaign_OR)*100, 1))`%, controlling for other variables
   
3. **Statistical significance**:
   - The effect is statistically significant (p < `r format.pval(summary(full_model)$coefficients["campaign", "Pr(>|z|)"], digits=4)`)
   - We can be 95% confident that the true odds ratio lies between `r round(campaign_CI[1], 4)` and `r round(campaign_CI[2], 4)`

So in a business perspective, even after accounting for demographic and financial characteristics:
- Additional campaign contacts are associated with lower subscription probability
- Each contact reduces the odds of subscription by approximately `r abs(round((1-campaign_OR)*100, 1))`%
- The negative relationship is consistent and statistically significant

## 1h. [10 points] Use the likelihood ratio test discussed in class to evaluate whether the model in 1f is a better fit than the model in 1d.

## 1h. Likelihood Ratio Test

Now I will compare the simpler model (1d) with campaign only to the full model (1f) with multiple predictors using the likelihood ratio test:

$$
LR = -2(\ln L_{\text{simple}} - \ln L_{\text{full}}) \sim \chi^2_{df}
$$

where $df$ is the difference in degrees of freedom between the models.

```{r}
# Performing likelihood ratio test
lr_test <- anova(logistic_model, full_model, test = "LRT")

# Creating detailed results table
lr_results <- data.frame(
  Model = c("Simple (1d)", "Full (1f)"),
  DF = c(lr_test$Df[1], lr_test$Df[2]),
  Deviance = c(lr_test$Deviance[1], lr_test$Deviance[2]),
  AIC = c(AIC(logistic_model), AIC(full_model)),
  BIC = c(BIC(logistic_model), BIC(full_model))
)

# Displaying model comparison statistics
knitr::kable(
  lr_results,
  caption = "Model Comparison Statistics",
  digits = 2,
  align = c('l', 'r', 'r', 'r', 'r')
)

# Creating likelihood ratio test results table
lr_stats <- data.frame(
  Metric = c("Likelihood Ratio Statistic", 
             "Degrees of Freedom", 
             "p-value",
             "Change in Deviance"),
  Value = c(
    lr_test$`LR stat`[2],
    lr_test$`Df`[2] - lr_test$`Df`[1],
    lr_test$`Pr(>Chi)`[2],
    lr_test$Deviance[1] - lr_test$Deviance[2]
  )
)

# Displaying likelihood ratio test results
knitr::kable(
  lr_stats,
  caption = "Likelihood Ratio Test Results",
  digits = 4,
  align = c('l', 'r')
)
```

Interpretting the results:

1. **Test Statistic**:
   - LR = `r round(lr_test$"LR stat"[2], 2)`
   - Degrees of freedom = `r lr_test$"Df"[2] - lr_test$"Df"[1]`
   - p-value < `r format.pval(lr_test$"Pr(>Chi)"[2], digits=4)`

2. **Model Comparison**:
   - Reduction in deviance: `r round(lr_test$Deviance[1] - lr_test$Deviance[2], 2)`
   - The full model has lower AIC (`r round(AIC(full_model), 2)` vs `r round(AIC(logistic_model), 2)`)
   - The full model has lower BIC (`r round(BIC(full_model), 2)` vs `r round(BIC(logistic_model), 2)`)

3. **Conclusion**: 
   The likelihood ratio test shows strong evidence (p < `r format.pval(lr_test$"Pr(>Chi)"[2], digits=4)`) that the full model provides a significantly better fit than the simple model. The substantial reduction in deviance and lower information criteria (AIC, BIC) further support using the more complex model for predicting term deposit subscriptions.

