---
title: "Homework 5: Namit Shrivastava"
format: pdf
editor: visual
---

## 1a. [10 points] Exploratory Analysis. Read the data into R. Report on each variable using a summary, a figure,  or a table as appropriate. You can ignore the channel(CHANNEL) and region (REGION) for this exercise.

First, let me read the data and perform exploratory analysis on the spending variables.

```{r}
# Loading the required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(GGally)
library(corrplot)

# Reading the data
wholesale_data <- read.csv("/Users/namomac/Downloads/Wholesale customers data.csv")

colnames(wholesale_data)

# Now the summary statistics for the numeric variables
numeric_vars <- wholesale_data[, c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")]
summary(numeric_vars)
```

Now let me show the distribution of Spending Variables
```{r}
# Creating histograms for each spending variable
spending_vars <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

# Creating individual histograms
hist_plots <- list()
for (var in spending_vars) {
  p <- ggplot(wholesale_data, aes_string(x = var)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
    theme_minimal() +
    labs(title = paste("Distribution of", var, "Spending"),
         x = paste(var, "(m.u.)"),
         y = "Frequency") +
    theme(plot.title = element_text(size = 10, face = "bold"))
  hist_plots[[var]] <- p
}

# Arranging histograms in a grid
grid.arrange(grobs = hist_plots, ncol = 2)
```

So when examining the summary statistics for the six spending categories, I noticed striking variations in customer purchasing patterns. 

1. The Fresh category showed the widest spending range meaning while some customers spent as little as 3 monetary units annually, one high-spending customer reached 112,151 units. However, the median (8,504 units) being significantly lower than the mean (12,000 units) suggests most customers cluster at lower spending levels with a few extreme outliers pulling the average up.

2. For Dairy products, I observed a similar pattern. So the average Milk spending (5,796 units) nearly doubled the median (3,627 units), indicating a right-skewed distribution. 

3. Grocery spending followed suit, with 75% of customers spending under 10,656 units despite the maximum reaching 92,780 units. 

4. The Frozen category surprised me with its particularly large gap between the median (1,526 units) and maximum (60,869 units), showing some clients make exceptionally large frozen goods purchases.


5. Detergents/Paper products stood out with the most dramatic disparity as the typical customer spent 816.5 units (median), the mean of 2,881.5 units revealed substantial high-end spenders. Delicatessen showed the tightest central tendency overall, though still contained extreme values up to 47,943 units.


Hence, across all categories, the consistent pattern of means exceeding medians and large gaps between third quartiles and maximums suggests most customers are moderate spenders, with a small but significant group of wholesale clients making exceptionally large purchases that distort the averages.


